---
title: "Assignment 2 - `dplyr` exploration"
author: "Lulu Pei"
date: "22/09/2019"
output: 
  html_document:
    toc: TRUE
    keep_md: TRUE
---

# Overview

Before playing around with the various `dplyr` commands, we must define a dataset. In this case, we will be looking at the `gapminder` dataset. To access the `gapminder` dataset, we need to load the `gapminder` package. To be able to call and use `dplyr` commands, we must load the `tidyverse` package as well. 

```{r load packages, warning=FALSE, message = FALSE}
library(gapminder)
library(tidyverse)
```

<br>

# Question 1 - Basic `dplyr`

## _1.1 - Filtering_

Filter `gapminder` data to contain only observations from Canada, the United States, and Mexico in the 1970s.

```{r, comment = NA}
gapminder %>%
  filter(country == 'Canada' | country == 'United States' | country == 'Mexico',
         year >= 1970 & year < 1980) %>%
  knitr::kable()
```

## _1.2 - Selecting_

Let's select only `country` and `gdpPercap` variables from our filtered subset of the `gapminder` dataset.

```{r, comment = NA}
gapminder %>%
  filter(country == 'Canada' | country == 'United States' | country == 'Mexico',
         year >= 1970 & year < 1980) %>%
  select(country, gdpPercap) %>%
  knitr::kable()
```

## _1.3 - Mutating_

Suppose we want to look at all the countries that have ever experienced a drop in life expectancy between 1952 and 2007. Let's define a new variable `lifeExp_change`, equaling the difference between life expectancy at one time point and life expectancy at the time point before.

```{r, comment = NA}
gapminder %>%
  group_by(country) %>%
  mutate(lifeExp_change = lifeExp - lag(lifeExp, order_by = year)) %>%
  filter(lifeExp_change < 0) %>%
  DT::datatable()
```

We can see that many countries have experienced a drop in life expectancy at some point between 1952 and 2007; however, what if we are only interested in countries that have experienced an overall life expectancy drop between the most recent year (2007) and the earliest year (1952).

```{r, comment = NA}
gapminder %>%
  filter(year == 1952 | year == 2007) %>%
  group_by(country) %>%
  mutate(lifeExp_change = lifeExp - lag(lifeExp)) %>%
  filter(lifeExp_change < 0) %>%
  knitr::kable()
```

From this output, we can see that the only countries that experienced an overall decline in life expectancy were Swaziland and Zimbabwe, with a decline of 1.794 and 4.964 years, respectively.

## _1.4 - Slicing_

Now, let's filter the `gapminder` dataset to show the maximum GDP per capita experienced by each country.

```{r, comment = NA}
gapminder %>%
  select(country, gdpPercap) %>%
  group_by(country) %>%
  slice(which.max(gdpPercap)) %>%
  rename(max_gdpPercap = gdpPercap) %>%
  DT::datatable()
```

This output allows us to determine in which year a country had its maximum GDP per capita, and what the life expectancy, population, and GDP per capita was at that time. 
<br>

## _1.5 - Plotting_

Let's investigate the relationship between life expectancy and GDP per capita in Canada. To do this, we will create a scatterplot using `ggplot2`. 

```{r fig.align="center"}
gapminder %>%
  filter(country == "Canada") %>%
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point(colour = "orange", size = 2) +
  scale_x_log10() +
  xlab("GDP per capita (log(US$))") +
  ylab("Life expectancy at birth (years)")
```

From this plot, we can see that in Canada, life expectancy has been increasing relatively linearly with the log transform of GDP per capita between the years of 1952 and 2007.

<br>

# Question 2 - Variable Exploration

To perform individual variable exploration using `dplyr` we will choose one categorical variable and one quantitative variable to explore. Let's say we want to analyze `continent` as the categorical variable and `population` as the quantitative variable.

## _2.1 - Categorical_ 

The categorical variable we are interested in exploring is `continent`. To start off, let's first investigate which continents are represented in our `gapminder` dataset. 

```{r, comment = NA}
levels(gapminder$continent)
```

Categorical variable exploration is usually performed through the generation of frequency tables - let's generate one for the `continent` variable. 

```{r}
gapminder %>%
  count(continent) %>%
  knitr::kable()
```

This command prints out the number of observations for each continent in our dataset. However, by looking at the `gapminder` dataset, we see that each country contributes 12 observations (represeting 12 time points) to the dataset. Suppose we want to know how many countries are in each continent for the `gapminder` dataset. To determine this, we need to remove replicates of the same country. The simpliest way to achieve this is to consider continent counts at each time point.

```{r}
gapminder %>%
  group_by(year) %>%
  count(continent) %>%
  DT::datatable()
```

We see that country distirbution does not change from year to year - there are the same number of countries in each of the 5 continents: Africa, the Americas, Asia, Europe, and Oceania. In our dataset, 52 countries are represented from Africa, 25 countries from the Americas, 33 countries from Asia, 30 countries from Europe, and 2 countries from Oceania. Let's view this data graphically using a bar graph - because continent counts do not change between years, it is sufficient to plot data from a single time point.

```{r, fig.align='center'}
gapminder %>%
  group_by(year) %>%
  filter(year==2007) %>%
  ggplot(aes(continent)) +
  geom_bar(fill = "orange", alpha = 0.7) +
  ylab("Number of Countries") +
  xlab("Continent")
```

## _2.2 - Quantitative_

Quantitative variable exploration is usually performed through the generation of 5-number summaries: min, 1st quartile, median, 3rd quartile, max. Let's generate a variation of the 5-number summary for the `pop` variable, including a measure of spread. 

```{r, comment = NA}
gapminder %>%
  summarise("min_pop (million)" = min(pop)/(10^6), "mean_pop (million)" = mean(pop)/(10^6), 
            "median_pop (million)" = median(pop)/(10^6), "max_pop (million)" = max(pop)/(10^6), 
            "sd_pop (million)" = sd(pop)/(10^6)) %>%
  knitr::kable()
```

From the output, we can observe the minimum and maximum values for population as well as the median and mean values and the standard deviation. We observe a deviation between population mean and population median, suggesting that the distribution is skewed with greater density towards lower values (`mean_pop` > `median_pop`).

However, these values are difficult to interpret because they refer to the population distribution across all countries between 1952 and 2007. It may be more informative if we investigated the change in population distribution from year to year, for example. 

```{r}
gapminder %>%
  select(year, pop) %>%  
  group_by(year) %>%
  summarise("min_pop (million)" = min(pop)/(10^6), "mean_pop (million)" = mean(pop)/(10^6), 
            "median_pop (million)" = median(pop)/(10^6), "max_pop (million)" = max(pop)/(10^6), 
            "sd_pop (million)" = sd(pop)/(10^6)) %>%
  knitr::kable()
```

From these summary statistics, we can see that there is a lot of variation in the `pop` variable (large standard deviation values). This is to be expected due to the large variation in population size across countries in the world. To visualize this graphically, we can plot side-by-side boxplots representing global population data at each time point.

```{r, fig.align='center'}
gapminder %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(year, pop)) +
  geom_boxplot() +
  scale_y_log10("Population (log scale)") +
  xlab("Year")
```

From this graph, we can observe a relatively consistent slow linear increase in log-transformed population over time, suggesting that between 1952 and 2007, global population has been steadily increasing. 

<br>

# Question 3 - Plot Exploration

## _3.1 - Scatterplot_

The first plot type that we are going to explore is the scatterplot. Suppose we want to group the `gapminder` dataset by continent, and then plot the relationship between life expectancy at birth and GDP per capita.

```{r fig.align='center'}
gapminder %>%
  group_by(continent) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, colour = continent)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  ylab("Life expectancy at birth (year)") +
  xlab("GDP per capita (log(US$))")
```

From the scatterplot, we can see that life expectancy and GDP per capita are roughly positively correlated, following a linear relationship. Not only does life expectancy tend to increase with GDP per capita, we can also observe that countries in Europe and Oceania tend to have higher life expectancy and GDP whereas countries in Africa tend to have lower life expectancy and GDP. Countries in the Americas and Asia seem to have more varied distributions of life expectancy and GDP with more discrepancy between countries.

## _3.2 - Time Series_

The second plot type we are going to explore is a time series. Suppose we want to investigate how population has changed over time between 1952 and 2007 in Canada, France, and Australia. We can visualize this through making a time series plot and fitting a trendline. 

```{r, fig.align='center'}
gapminder %>%
  filter(country == "Canada" | country == "France" | country == "Australia") %>%
  ggplot(aes(x = year, y = pop, col = country)) +
  geom_point(size = 2) +
  geom_line(size = 1) +
  xlab("Year") +
  ylab("Population") +
  theme(legend.title=element_blank())
```

From observing the trendlines, we can observe that all countries have experienced roughly linear increases in population over time. Canada and France seem to have similar rates of population increase, while the rate of increase in population is slower in Australia. Between 1952 and 2007, Canada and France have experienced an increase in population of approximately 20 million people, while Australia's population has increased by approximately 10 million people. 

# Bonus - Recycling

Evaluation of the following command:

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan")) %>%
  DT::datatable()
```

The analyst's goal was to obtain a subset of the `gapminder` dataset, containing data from Rwanda and Afghanistan. However, the output of the above command only returns 12 observations when there should have been 24 (12 observations for each country representing the 12 time points between 1952 and 2007). Each year should have both a Rwanda observation and an Afghanistan observation; however, in our subset, each year is only represented by one country. This suggests that the subsetted data is incomplete. 

To correctly obtain all data from Rwanda and Afghanistan, we can use the logical "or" operator (denoted as "|" in `dplyr`). The code below will return a dataframe that contains all observations from either Rwanda or Afghanistan.

```{r}
filter(gapminder, country == "Rwanda" | country == "Afghanistan") %>%
  DT::datatable()
```

From this output, we see that we are left with a dataframe of the correct dimensions (24 observations) with all data from Rwanda and Afghanistan between 1952 and 2007. 
