---
title: "Gapminder Exploration"
author: "Lulu Pei"
date: "08/09/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
library(gapminder)
```

## Presentation Outline

<br>

1. Overview of `gapminder` dataset

2. Analysis of continuous variables

3. Analysis of categorical variables


## Gapminder Dataset

The `gapminder` dataset is a data frame consisting of `r nrow(gapminder)` observations and `r ncol(gapminder)` variables of interest. 

```{r, echo = TRUE, comment = NA}
head(gapminder)
```


***
The `r ncol(gapminder)` variables of interest are:

```{r, comment = NA}
colnames(gapminder)
```

<br>

#### Continuous variables: 
- `lifeExp` 
- `pop`
- `gdpPercap`

#### Categorical variables: 
- `country`
- `continent`
- `year`


## Continuous Analysis

Summarize continuous variables using **5-number summary**.

<br>

```{r, comment = NA}
summary(gapminder[,4:6])
```


***
### `lifeExp` variable
- `mean` approximately equal to `median`
- symmetric distribution

### `pop` variable
- `mean` >> `median`
- positively skewed distribution 

### `gdpPercap` variable
- `mean` > `median`
- positively skewed distribution


## Categorical Analysis

Summarize categorical variables using **contingency table**.

<br>

- Each year point is represented 142 times in the dataset
  - Reflects the 142 countries surveyed

<br>
  
- Each country is represented 12 times in the dataset
  - Reflects the 12 year points at which population data was collected 

***
```{r, comment = NA, echo = TRUE}
table(gapminder$continent)
```

- Frequencies represent total number of continent occurrences
- However, each country contributes 12 observations from different years
- Continent distribution can be determined by removing country duplicates

```{r, comment = NA, echo = TRUE}
table(gapminder$continent) / 12
```


## Summary

`Gapminder` dataset contains population data from 52 countries in Africa, 25 countries in the Americas, 33 countries in Asia, 30 countries in Europe, and 2 countries in Oceania, representing a total of 142 countries globally. 

<br>

Distribution of life expectancy is roughly symmetric, while GDP per capita and population distributions are positively skewed.
