---
title: "Gapminder Exploration"
author: "Lulu Pei"
date: "12/09/2019"
output: 
  github_document:
    toc: TRUE
---

## _Overview_

In this document, we will be performing some basic data exploration using the `gapminder` dataset.

First, let's load packages. We will need to load the `gapminder` package to access the `gapminder` data set.

```{r} 
library(gapminder)
```

What exactly does the `gapminder` dataset contain? Let's find out by taking a look at the dataset description excerpt provided by R.

```{r eval = FALSE, comment = NA}
?gapminder
```

From this excerpt we can see that the dataset contains yearly population, life expectancy, and GDP per capita data for 142 countries representing 5 continents.

Here are the first 10 lines of the dataset to present an overview:

```{r, comment = NA}
head(gapminder, 10)
```

Prior to performing any specific analysis, let's obtain some summary statistics for the variables contained in this dataset. 

```{r, comment = NA}
summary(gapminder)
```

From the above output, we can read off 5-number summaries for the continuous variables - `lifeExp`, `pop`, and `gdpPercap` - and frequency data for the categorical variables - `country`, `continent`, and `year`. 

With this basic understanding of the type of data that is contained within this dataframe, let's perform some exploratory data analysis!


## _Subsetting Dataset_

Through observing the dataset, we can see that each country has population data ranging from 1952 to 2007 in increments of 5 years. This is a large dataset, so let's breakdown the type of data that we are dealing with.

```{r, comment= NA}
str(gapminder)
```

From the output of this command, we can see that the `gapminder` dataset contains 6 variables - three of which (`country`, `continent`, and `year`) are categorical, and three of which (`lifeExp`, `pop`, `gdpPercap`) are continuous. Suppose we are interested in investigating how `lifeExp`, `pop` and `gdpPercap` variables have changed over time in Canada. For simplicity, let's select three time points: 1952, 1977, and 2007.

We will define dataframes `past` to contain all observations from 1952, `mid` to contain all observations from 1977, and `present` to contain all observations from 2007.

```{r}
past <- subset(gapminder, year==1952)

mid <- subset(gapminder, year==1977)

present <- subset(gapminder, year==2007)
```


## _Analysis_

Now that we have all the prep work done, let's begin our analysis! Starting with 1952, what were the population characteristics in Canada? 

```{r, comment=NA}
past[which(past$country=="Canada"), 4:6]
```

From this output we can observe that in 1952, Canada had a life expectancy at birth of 68.8 years, a population of 14,785,584 people, and a GDP per capita of US$11,367. 

Now, let's take a look at 1977. 

```{r, comment=NA}
mid[which(mid$country=="Canada"), 4:6]
```

In 1977, Canada had a life expectancy at birth of 74.2 years, a population of 23,796,400 people, and a GDP per capita of US$22,091.

Finally, let's analyze the most recent time point, 2007. 

```{r, comment=NA}
present[which(present$country=="Canada"), 4:6]
```

Using the same technique, we observe that in 2007, Canada had a population of 33,390,141 people, with a life expectancy at birth of 80.7 years and a GDP per capita of US$36,319.

## _Conclusion_

Through this basic analysis of the Canadian subset of the `gapminder` dataset, we can infer that the population characteristics of Canada have been increasing relatively linearly between the years of 1952 and 2007. Population increases by approximately 10 million people every 20 years, while life expectancy increases by 6 years and GDP per capita increases by approximately US$11,000.
