---
title: "Gapminder Exploration"
author: "Lulu Pei"
date: "08/09/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

## _Overview_

In this document, we will be performing some basic data exploration using the `gapminder` dataset. 

First, let's load packages. We will need to load the `gapminder` package to access the `gapminder` data set.

```{r} 
library(gapminder)
```

What exactly does the `gapminder` dataset contain? Let's find out by taking a look at the dataset description excerpt provided by R.

```{r eval = FALSE}
?gapminder
```

From this excerpt we can see that the dataset contains yearly population, life expectancy, and GDP per capita data for 142 countries representing 5 continents.

Here are the first 10 lines of the dataset to present an overview:

```{r}
head(gapminder, 10)
```

With this basic understanding of the type of data that is contained within this dataframe, let's perform some exploratory data analysis!


## _Analysis_

Through observing the dataset, we can see that each country has population data ranging from 1952 to 2007 in increments of 5 years. This is a large dataset, so let's breakdown the type of data that we are dealing with.

```{r}
str(gapminder)
```

From the output of this command, we can see that the `gapminder` dataset contains 6 variables - three of which (`country`, `continent`, and `year`) are categorical, and three of which (`lifeExp`, `pop`, `gdpPercap`) are continuous. To obtain summaries of each variable, we must consider the categorical and continuous variables separately. 

Let's first consider the continuous variables: `lifeExp`, `pop`, and `gdpPercap`. The most efficient summary to obtain for continuous variables is the 5-number summary, providing information on the mean, median, minimum and maximum values, as well as quantile values. 

```{r}
summary(gapminder[,4:6])
```

From the 5-number summaries of each variable, we can make some inferences about their distribution. Because the `mean` and `median` for `lifeExp` are approximately equal, we can conclude that the distribution for `lifeExp` is roughly symmetric and normally distributed. Regarding both the `pop` and `gdpPercap` variables, the `mean` is greater than the `median`, suggesting that the distributions of these variables are positively skewed with larger values being more spread out than smaller values. 

Now, let's consider the categorical variables: `country`, `continent`, and `year`. The most efficient way to summarize such variables is through the use of contingency tables. 

```{r}
table(gapminder$country)

table(gapminder$continent)

table(gapminder$year)
```

From the output, we can see that each country is represented 12 times in the dataset, which reflects the 12 year points at which data was collected. Each year point is represented 142 times in the dataset, reflecting the 142 countries included in the `gapminder` dataset. We can also observe the country distribution by continent. To do this, we must take `continent` frequencies and divide by the number of year points to remove duplicating countries. 

```{r}
table(gapminder$continent) / 12
```

This command allows us to conclude that the `gapminder` dataset contains population data from 52 countries in Africa, 25 countries in the Americas, 33 countries in Asia, 30 countries in Europe, and 2 countries in Oceania, representing a total of 142 countries globally. 
