---
title: "Gapminder Exploration"
author: "Lulu Pei"
date: "12/09/2019"
output: ioslides_presentation
---

```{r, include=FALSE}
library(gapminder)
```

## Presentation Outline

<br>

1. Overview of `gapminder` dataset

2. Subsetting dataset

3. Analysis of population characteristics

4. Conclusion


## Gapminder Dataset

The `gapminder` dataset is a data frame consisting of `r nrow(gapminder)` observations and `r ncol(gapminder)` variables of interest. 

```{r, echo = TRUE, comment = NA}
head(gapminder)
```


***
The `r ncol(gapminder)` variables of interest are:

```{r, comment = NA}
colnames(gapminder)
```

<br>

Continuous Variables | Categorical Variables
-------------------- | ---------------------
`lifeExp` | **`country`**
`pop` | **`continent`**
`gdpPercap` | **`year`**


## Summary Statistics

```{r, comment = NA}
summary(gapminder)
```


## Subsetting Dataset

**Research Question**: How have `lifeExp`, `pop` and `gdpPercap` variables changed over time in Canada?

For simplicity, consider three time points: 1952, 1977, 2007

Define dataframes `past` to contain all observations from 1952, `mid` to contain all observations from 1977, and `present` to contain all observations from 2007.

```{r}
past <- subset(gapminder, year==1952)

mid <- subset(gapminder, year==1977)

present <- subset(gapminder, year==2007)
```

## Analysis

Population characteristics of Canada in **1952**:

```{r, comment=NA}
past[which(past$country=="Canada"), 4:6]
```

Life expectancy at birth = 68.8 years

Population = 14,785,584 people

GDP per capita = US$11,367

***

<br>

Population characteristics of Canada in **1977**:

```{r, comment=NA}
mid[which(mid$country=="Canada"), 4:6]
```

Life expectancy at birth = 74.2 years

Population = 23,796,400 people

GDP per capita = US$22,091

***

<br>

Population characteristics of Canada in **2007**:

```{r, comment=NA}
present[which(present$country=="Canada"), 4:6]
```

Life expectancy at birth = 80.7 years

Population = 33,390,141 people

GDP per capita = US$36,319


## Conclusion

Population characteristics of Canada have been increasing relatively linearly between the years of 1952 and 2007.

<br>

Population increases by approximately 10 million people every 20 years, while life expectancy increases by roughly 6 years and GDP per capita increases by approximately US$11,000.
